Inhibition of O-GlcNAc transferase activates type I interferon-dependent antitumor immunity by bridging cGAS-STING pathway

Version 1

Abstract

Abstract

The O-GlcNAc transferase (OGT) is an essential enzyme that mediates protein O-GlcNAcylation, a unique form of posttranslational modification of many nuclear and cytosolic proteins. Recent studies observed increased OGT and O-GlcNAcylation levels in a broad range of human cancer tissues compared to adjacent normal tissues, indicating a universal effect of OGT in promoting tumorigenesis. Here, we show that OGT is essential for tumor growth in immunocompetent hosts by repressing the cyclic GMP-AMP synthase (cGAS)-dependent DNA sensing pathway. We found that deletion of OGT (Ogt−/−) caused a marked reduction in tumor growth in both syngeneic tumor models and a genetic colorectal cancer (CRC) model induced by mutation of the Apc gene (Apcmin). Pharmacological inhibition or genetic deletion of OGT induced a robust genomic instability (GIN), leading to cGAS-dependent production of the type I interferon (IFN-I) and IFN-stimulated genes (ISGs). As a result, deletion of Cgas or Sting from Ogt−/− cancer cells restored tumor growth, and this correlated with impaired CD8+ T cell-mediated antitumor immunity. Mechanistically, we found that OGT-dependent cleavage of host cell factor C1 (HCF-1) is required for the avoidance of GIN and IFN-I production in tumors. In summary, our results identify OGT-mediated genomic stability and activate cGAS-STING pathway as an important tumor cell-intrinsic mechanism to repress antitumor immunity.

Introduction

Cancer cells can maintain malignant phenotypes partially due to altering the post-translational modification (PTM) patterns of cancer-related functional proteins under the stimulation of extracellular and intracellular factors [1]. Protein modification by the O-linked β-N-acetylglucosamine (O-GlcNAc) is a dynamic and reversible post-translational modification, which is added to the hydroxyl group of a specific serine or threonine residue in a target protein by O-GlcNAc transferase (OGT) and removed by O-GlcNAcase (OGA) [2]. O-GlcNAcylation is a fast-cycle and nutrient-sensitive PTM, which modifies thousands of cytoplasmic, nuclear, and mitochondrial proteins and mediates crosstalk with protein phosphorylation, regulating signal transduction and affecting protein localization, activity, stability, and protein-protein interaction. Dysregulation of O-GlcNAcylation is associated with multiple metabolic diseases and cancer [3, 4].

Recent studies observed increased OGT and O-GlcNAcylation level in human colon cancer tissues compared to adjacent normal tissues [5–7], indicating an essential role of OGT-mediated protein O-GlcNAcylation in the pathogenesis of colon cancer. Several oncogenic proteins that are involved in the pathogenesis of colon cancer have been shown to be directly modified by O-GlcNAc, including β-catenin and NF-κB [8–10]. For example, O-GlcNAcylation of β-catenin at T41 inhibits its phosphorylation, which subsequently attenuates its ubiquitination and degradation and promotes oncogenic activity [7, 11]. In contrast, inhibitory roles of O-GlcNAc signaling in the growth of human colon cancer and in oncogenic Wnt/β-catenin signaling have also been reported [12, 13]. The aforementioned studies show that O-GlcNAc modification of specific proteins can play opposing roles in tumorigenesis. However, the overall effect of OGT-mediated O-GlcNAcylation in cancer remains unknown.

The cGAS/STING cytosolic DNA-sensing pathway plays a vital role in activating the innate immune response and production of the type I interferons (IFN-I) [14]. Cyclic guanosine monophosphate (GMP)–adenosine monophosphate (AMP) synthase (cGAS) interacts with cytosolic double-stranded DNA (dsDNA) in a sequence-independent manner. The direct binding of cGAS to cytosolic dsDNA promotes cGAS homodimerization and activates the catalytic activity of cGAS, producing 2’, 3’-cyclic GMP-AMP (cGAMP) from ATP and GTP. The second messenger cGAMP binds to and activates the endoplasmic resident stimulator of interferon genes (STING). Once activated, STING translocates to the ER-Golgi intermediate compartment (ERGIC) to recruit TANK-binding kinase 1 (TBK1) and IFN regulatory factor 3 (IRF3), leading to the production of IFN-I and activation of numerous IFN-stimulated genes (ISGs) [15–17]. However, it remains largely unknown whether OGT expression affects cGAS-STING pathway and antitumor immunity.

In this study, we find that deficiency or pharmacological inhibition of OGT and subsequent accumulation of cytosolic dsDNA activates the cGAS-STING pathway and induces CD8+ T cell-dependent antitumor immunity. Deletion of cGAS or STING diminishes DNA sensing and lead to progressive tumor growth. Mechanistically, we show that OGT could interacts with HCF-1 and cleaves it, which contributes to the maintenance of genomic stability. Re-expression of HCF-1C600 in Ogt−/− tumor cells inhibit production of cytosolic dsDNA and IFN-I. In summary, our findings demonstrate that OGT-mediated DNA damage and activate cGAS-STING pathway as an important tumor cell-intrinsic mechanism to repress antitumor immunity and provides a window for potential therapeutic opportunities for in OGT-dependent cancer.

Results

Increased OGT expression in human and mouse tumor samples

Protein O-GlcNAcylation is upregulated in various cancers [12, 18–21]. OGT is the only known enzyme that mediates O-GlcNAcylation of proteins at the Ser or Thr residues [22], we hypothesized that OGT could serve as an important regulator to regulate cancer cell growth and serve as a biomarker for cancer. Initial analysis of data from The Cancer Genome Atlas (TCGA) dataset in GEPIA2 (http://gepia2.cancer-pku.cn/#index) and found a significant positive correlation between OGT mRNA expression and tumorigenesis in bladder urothelial carcinoma (BLCA), cholangiocarcinoma (CHOL), colon adenocarcinoma (COAD), esophageal carcinoma (ESCA), head and neck squamous cell carcinoma (HNSC), kidney chromophobe (LIHC), lung adenocarcinoma (LUAD), prostate adenocarcinoma (PRAD), rectum adenocarcinoma (READ), sarcoma (SARC), and a stomach adenocarcinoma (STAD) (Supplemental Fig. 1A). Next, we utilized UALCAN database analysis, and found that OGT mRNA expression was compared between 286 COAD samples and 41 adjacent or normal samples, the expression of OGT was significantly increased in COAD at the transcriptional level (Fig. 1A). A significant positive correlation tendency between OGT mRNA expression and individual different tumor stages was observed. The stage IV COAD tissues exhibited the highest expression level of OGT in compared with low stage (Fig. 1B). We also observed a significant positive correlation between OGT mRNA expression and nodal metastasis status (Fig. 1C). Furthermore, we found that the protein level of OGT was also significantly increased in COAD patient samples and amongst individual different stages based on the CPTAC and HPA online database (Fig. 1D-F). Similar results are found in LUAD (Supplemental Fig. 1B-F). These results suggest that high mRNA and protein levels of OGT in tumorigenesis were consistent in different databases.

OGT is significantly upregulated in human and mouse tumor samples.

A-C) Boxplot showing mRNA expression level of Ogt, Normal and tumor samples (A), Individual stages (B), Nodal metastasis status (C). The plot was generated using the UALCAN online server. D-E) Boxplot showing protein expression level of OGT, Normal and tumor samples (A), Individual stages (B). The plot was generated using the UALCAN online server (https://ualcan.path.uab.edu/analysis.html). F) IHC analysis of OGT expression in normal colon tissues, primary colon tumor samples (from Human Protein Atlas, https://www.proteinatlas.org/), scale bar: 400μm. G) Western blot analysis of OGT and O-GlcNAc in normal, adjacent and tumor tissues in Apcminspontaneous tumor mice. H) HE staining and IHC staining of OGT in adjacent and tumor tissues in Apcmin spontaneous tumor mice, scale bar: left panel 275 μm, right panel 75μm, n=3 respectively. I) Schematic of AOM/DSS model of colitis-associated colorectal cancer (CAC). J) Western blot analysis of OGT and O-GlcNAc in normal, adjacent and tumor tissues in CAC model. K) HE staining and IHC staining of OGT in adjacent and tumor tissues in CAC model, scale bar: left panel 275 μm, right panel 75μm, n=4 respectively. human samples (A-F), mouse samples (G-K). Statistical significance was determined by Pearson test, unpaired Student’s t-test, *p < 0.05, **p < 0.01, ***p < 0.001, ns, no significant difference. Data represent the mean of ± SD.

To determine whether intestinal OGT expression was increased in Apcmin colorectal tumor mouse model, intestinal tissues were collected for western blot and immunohistochemical staining (IHC) analysis. Intestinal OGT protein was markedly increased in Apcmin mouse tumor tissues compared to adjacent or normal tissues (Fig. 1G). As expected, O-GlcNAcylation proteins levels in intestinal tissues were also significantly higher in tumor tissues than adjacent or normal tissues. IHC staining revealed that mouse OGT protein was markedly higher in tumor tissues than in adjacent tissues (Fig. 1H), which was consistent with the expression pattern in human samples. OGT protein was also upregulated in the azoxymethane (AOM)/dextran sodium sulfate (DSS) colorectal tumor model (Fig. 1I-K). Together, these data strongly suggest that OGT may play a critical role in tumorigenesis and serve as a prognostic marker and therapeutic target in cancer treatment.

Epithelial OGT deletion inhibits mouse colorectal tumorigenesis

To determine whether OGT may be an important therapeutic target for tumor treatment, we generated an intestinal epithelial cell-specific Ogt deletion (OgtΔIEC) by crossing Ogtfl/fl mice with Villin-Cre mice. OgtΔIEC mice were then crossed with Apcmin mice to generate OgtΔIECApcmin mice (Supplemental Fig. 2). To determine whether intestinal OGT expression was decreased in OgtΔIECApcmin mice, intestinal tissues were collected for western blot analysis. Intestinal OGT levels were drastically decreased in OgtΔIECApcmin mice compared to wildtype Apcmin mice (Fig. 2A). Importantly, deletion of the Ogt in intestinal epithelial cells resulted in significantly reduced tumor size and total number of polyps at 20 weeks of age (Fig. 2A), indicating that OGT promotes oncogenic transformation in colorectal tumor in vivo. To gain insight into the role of protein on intestinal carcinogenesis in intestinal cancer, we examined the differences in intestinal carcinogenesis. Hematoxylin and eosin (H&E) staining demonstrated that epithelial inflammation diminished in OgtΔIECApcmin mice compared to wildtype Apcmin mice (Fig. 2B). OGT deficiency in intestinal tissues was also associated with significantly elevated gene expression of key pro-inflammatory cytokines like interleukin Il1α, Il6, TNF-α, as well as several interferons and ISGs like Isg15, Mx1, Cxcl10 (Fig. 2C). Furthermore, we found IL-2, IL-6, IL-10, IL-17, IFN-α, IFN-β, IFN-γ and CXCL10 were upregulated compared with control mice (Fig. 2D).

Epithelial OGT deletion inhibits mouse colorectal tumorigenesis.

A) Western blot analysis of OGT expression and tumor number in Ogt△IEC model. B) HE staining, scale bar: up panel 275 μm, bottom panel 20 μm, n=3 respectively. C) Real-time PCR analysis of mRNA expression in intestine. D) ELISA analysis in intestine. Statistical significance was determined by unpaired Student’s t-test, *p < 0.05, **p < 0.01, ***p < 0.001, ns, no significant difference. Data represent the mean of ±SD.

OGT deficiency activates cGAS/STING-dependent IFN-I pathway

Based on previous studies, we found that several interferons and several interferon-stimulated genes, such as Isg15, Mx1, IFN-α, IFN-β and Cxcl10 were upregulated. We hypothesized that OGT deficiency may activate the type I IFN pathway. First, we generated Ogt knock outs of murine colorectal carcinoma models like MC38, LLC, B16-OVA and of a human colorectal cell line, HT29. We then evaluated the expression changes of Ifna4, Ifnb1, Isg15, Mx1 and found that all the genes except Ifna4 mRNA expression were significantly increased in all cultured Ogt knockout cells (Fig. 3A-D). Next, we investigated how the type I IFN pathway in Ogt knockout cells was activated [23, 24]. We found that while the phosphorylation of STAT1, TBK1 and IRF3 is increased, the STING expression is reduced in Ogt knockout MC38, LLC, B16-OVA and HT29 cells (Fig. 3E-F). To determine whether this effect is specific for Ogt knockout, we stably expressed exogenous OGT in Ogt knockout cells and found that the type I IFN pathway activation effect is abolished (Fig. 3F). The type I IFN pathway is typically activated by both viral RNA and dsDNA. In order to eliminate the type I IFN pathway activate independent on RNA but dependent on dsDNA, we knocked out MAVS protein in Ogt−/− cells, which is a RNA pivotal adaptor protein activate the downstream protein kinase TBK1, IFN pathway and ISGs in Ogt−/− cells (24), and we found that knockout of MAVS in Ogt−/− cells has no effect on either Isg15, Mx1, Ifna4 and Ifnb1 expression or the type I IFN pathway (Fig. 3G-H).

OGT deficiency induces cGAS/STING-dependent the type I IFN pathway.

A-D) Real-time PCR analysis of mRNA expression in different Ogt knockout gRNA clones in MC38 (A), LLC (B), HT29 (C), B16-OVA (D) cells. E) Western blot analysis of protein expression in different Ogt−/− clones in MC38, LLC and B16-OVA cells. F) Western blot analysis of protein expression in Ogt−/− rescued MC38, LLC, HT29 and B16-OVA cells. G-H) Real-time PCR and western blot analysis of mRNA and protein expression in different Ogt−/−Mavs−/− double knockout clones in MC38 cells. I-K) Real-time PCR and ELISA analysis of expression in Ogt−/−cGAS−/− double knockout clones in MC38 (I), MC38 (J), HT29 (K) cells. L) Western blot analysis of protein expression in Ogt−/−cGAS−/− or Ogt−/−Sting−/− double knockout clones in MC38, HT29 and B16-OVA cells. M-N) BMDCs pre-treated with B16-OVA-Ogt−/− (L), B16-OVA-Ogt−/−cGAS−/− or B16-OVA-Ogt−/−Sting−/− double knockout (M) supernatant, and co-cultured with OT-1 T cell, then T cell proliferation was evaluated by flow cytometry, OVA257-264 as a positive control. Representative fluorescence-activated cell sorting histograms and statistical data are shown. Data are representative of two or three independent experiments. Statistical significance was determined by unpaired Student’s t-test, one-way ANOVA, two-way ANOVA, *p < 0.05,**p < 0.01,***p < 0.001, ns, no significant difference. Data represent the mean of ±SD.

We next investigated the mechanism of STING activation, and found that Ogt knockout activate the cGAS pathway which degrades the STING. This result is similar to previous studies [25, 26]. The phenotype of Ogt knockout implies that STING is involved in the activation of the type I IFN pathway. We hypothesized that OGT negatively regulates type I IFN through the cGAS/STING pathway. Next, we produced the OGT/cGAS and OGT/STING double knockout cells both in MC38, HT29 and B16-OVA. Surprisingly, these type I IFN and Cxcl10 signals disappeared in OGT/cGAS and OGT/STING double knockout cells both in mRNA and protein level (Fig. 3I-L).

The type I IFN signaling is a key pathway that promotes antigen presentation and DC activation [27–29]. Phagocytosis of extracellular tumor DNA by DCs triggers the activation of the cGAS-STING-IFN pathway [30, 31]. To determine whether the increased cGAS-STING-IFN signaling in Ogt knockout cells provides an activated signal for single epitope-specific CD8+ T cell priming by antigen-presenting cells (APCs), we added the supernatant from B16-OVA tumor cells into the co-culture system of BMDCs and OT-I cells. The results showed that BMDCs pre-cultured with Ogt+/+ or Ogt−/− B16-OVA tumor cells supernatant provided a potent activated signal for optimal single epitope-specific T cell proliferation (Fig. 3M), while cGAS or STING deficiency in Ogt−/− tumor cells diminished such effects (Fig. 3N). Further, we added the supernatant from Ogt+/+ or Ogt−/− B16-OVA tumor cells into the co-culture system of Ifnar1 knockout BMDCs and OT-I cells. The cell proliferation was abolished when treated with Ifnar1-/- BMDCs pre-cultured with Ogt+/+ or Ogt−/− B16-OVA tumor cells supernatant. (Supplemental Fig. 3). These results demonstrated that Ogt deficiency activate antitumor CD8+ T cells response is dependent on the type I IFN signal in dendritic cells and in a manner dependent on the tumor-cell-intrinsic cGAS-STING pathway.

OGT deficiency causes DNA damage and cytosolic DNA accumulation

Because the accumulation of cytosolic DNA is a consequence of nuclear DNA damage, which can activate immune response. The cGAS senses cytoplasmic DNA as a consequence of nuclear DNA damage [32]. Previous studies showed that the STING is degraded while cGAS-STING is activated [25]. As mentioned above, the STING degraded in Ogt knockout cells. We hypothesized that whether the cytosolic DNA is accumulated in Ogt knockout cells, we stained cytosolic double-strand DNA (dsDNA) with PicoGreen, a widely used immunofluorescence staining that selectively binds to dsDNA [33, 34], and found that a significantly higher percentage of Ogt deficiency cells than control cells in two different Ogt knockout MC38 and LLC cell clones, respectively (Fig. 4A and Supplemental Fig. 4A). In addition, we also stained dsDNA with anti-dsDNA immunofluorescence and the results were similar to PicoGreen staining (Fig. 4B and Supplemental Fig. 4B). We next assessed the phosphorylation of H2AX at Ser 139 (γH2AX), an indirect marker of DNA DSBs in the cell lines [35]. Comparable expression levels of H2AX, we found that γH2AX was dramatically increased in Ogt knockout MC38, LLC, HT29 and B16-OVA cells (Fig. 4C and 4F). We also used the anti-γH2AX immunofluorescence staining and comet assays, a classical quantifying and analyzing DNA damage, and found that Ogt knockout significantly induced immense DNA strand breakage in MC38 cells (Fig. 4D-E). Finally, the rescued results showed that exogenous OGT expression abolished the γH2AX expression both in MC38, LLC, HT29 and B16-OVA cells and DNA damage in MC38 cells (Fig. 4F-H). Together, these data indicated that Ogt knockout caused DNA damage and induced the cytosolic DNA accumulation.

OGT deficiency causes DNA damage and accumulates cytosolic DNA.

A) The dsDNA in different Ogt−/− clones were determined by PicoGreen staining assay and extranuclear dsDNA was quantified by image J in MC38 cells. B) The dsDNA in different Ogt−/− clones were determined by anti-dsDNA fluorescence staining assay and extranuclear dsDNA was quantified by image J in MC38 cells. C) Western blot analysis of γH2AX and H2AX in different Ogt−/− clones in MC38, LLC and B16-OVA cells. D) The anti-γH2AX staining assay in different Ogt−/− clones and extranuclear dsDNA was quantified by image J. E) The DNA damage was determined by comet assay, and extranuclear dsDNA was analyzed by using CometScore. F) Western blot analysis of γH2AX and H2AX in Ogt−/− rescued clone in MC38, LLC, HT29 and B16-OVA cells. G) The DNA damage in rescued MC38 cells were determined by comet assay, and extranuclear dsDNA was analyzed by using CometScore. Data are representative of three or four independent experiments. Statistical significance was determined by unpaired Student’s t-test, one-way ANOVA, *p < 0.05, **p < 0.01, ***p < 0.001, ns, no significant difference. Data represent the mean of ±SD.

The C-terminus of HCF-1 rescues DNA damage and IFN-I pathway in Ogt−/− cells

Previous studies in Fig. 3F and Fig. 4F showed that expression of exogenous OGT abolished the type I IFN and γH2AX signals. To explore the mechanism how OGT regulate the IFN and γH2AX production. Next, we use GFP agarose immunoprecipitation and liquid chromatography coupled to tandem MS (LC-MS/MS) to identify different proteins that interacted with OGT in human colonic cancer cell line HT29. Interestingly, HCF-1 was the most enriched protein in the precipitates from OGT restored HT29 knockout cells compared to OGT knockout cells reconstituted with empty vector, based on the number of peptides (indicating the identification confidence) and the number of peptide-spectrum matches (PSMs, indicating the abundance) (Fig. 5A and Supplemental Table 4). Previous studies showed that the human epigenetic cell-cycle regulator (HCF-1) undergoes an unusual proteolytic maturation by OGT cleavage and process resulting in stably associated HCF-1N1011 and HCF-1C600 subunits that regulate different aspects of the cell cycle [36]. We confirmed these results by using co-immunoprecipitation assay (Fig. 5B-C). We rescued the HCF-1 cleavage phenotype by expression of exogenous OGT (Fig. 5D). To clarify how to OGT regulate function of HCF-1, we transfected empty vector (EV), the full-length HCF-1-HA (HCF-1FL), HCF-11-1011-HA (HCF-1N1011), HCF-11-450-HA (HCF-1N450), HCF-1450-1011-HA (HCF-1N450-1011) and HCF-11436-2035-HA (HCF-1C600) and myc-OGT and found that OGT can bind HCF-1C600 (Fig. 5E), furthermore OGT directly bind the HCF-1C600 (Supplemental Fig. 5A-B). To further understand the physiological function of HCF-1, we transfected EV, HCF-1FL, HCF-1N1011and HCF-1C600 into MC38 control and OGT knockout cells, respectively. As shown in Fig. 5F, transfected HCF-1FL and HCF-1N1011 didn’t rescue the gene expression differences, such as Isg15, Mx1 and Ifnb1, but HCF-1C600 can restore the gene expression differences. These results were confirmed by using western blot and anti-dsDNA immunofluorescence staining (Fig. 5G-H). Overall, these data indicated that OGT regulates the HCF-1 cleavage and maturation, HCF-1C600 can eliminate the cytosolic DNA accumulation, DNA damage, the type I IFN activation and restrain cGAS-STING-mediated DNA sensing.

The C terminal of HCF-1 rescue DNA damage and the type I IFN pathway in Ogt−/− cells.

A) Volcano plot of proteins identified by LC–MS/MS in stably expressed exogenous OGT in OGT knockout HT29 cells. B) Immunoprecipitation and western blot of OGT knockout HT29 rescued cells. C) Immunoprecipitation and western blot of 293T cells. D) western blot of Ogt knockout MC38 rescued cells. E) Co-IP analysis of the interaction between OGT and different HCF-1 mutant. F) Real-time PCR analysis of mRNA expression in different HCF-1 isoforms in MC38 OGT knockout cells. G) Western blot analysis of protein expression in exogenous HCF-1C600 expression in MC38 Ogt knockout cells. H) The anti-dsDNA fluorescence staining in exogenous HCF-1C600 expression in MC38 OGT knockout cells. Data are representative of three or four independent experiments. Statistical significance was determined by one-way ANOVA, two-way ANOVA, *p < 0.05, **p < 0.01, ***p < 0.001, ns, no significant difference. Data represent the mean of ±SD.

OGT deficiency inhibits tumor progression through enhancing infiltration by CD8+ T cells

We examined whether inhibiting Ogt can delay the tumor growth and prolong the survival. To further verify these effects, we utilized transplanted tumor model. We found that the Ogt deficiency MC38 colorectal tumor cells have no obvious inhibit cell growth in vitro (Supplemental Fig. 6A). However, we found that a significant delay in tumor growth, tumor weight and prolong mice survival compared to the control group (Fig. 6A-B). We also used lewis lung carcinoma (LLC) cells and the B16-OVA melanoma cells because they represent an aggressive murine tumor model and are highly resistant to various immunotherapies, similar results were shown in Fig. 6C-D and Supplemental Fig. 6B-E. These results showed that Ogt deficiency could delay both tumor growth and prolong survival of MC38, LLC, B16-OVA in tumor model. We next examined whether the enhanced antitumor by utilizing tumor microenvironment (TME). Consistently, tumors from C57BL/6 immunocompetent mice bearing Ogt−/− MC38 tumors compared with MC38 control tumors, showed higher proportion of CD8+ and CD4+ T cells, functional CD8+ IFN-γ+, CD8+ TNF-α+ and CD8+ IFN-γ+ TNF-α+ double positive T cells, but not CD45+, CD11b+ CD11c+, CD11b+ F4/80+, CD11b+ Ly6C+ and Treg cells (Fig. 6E-H and Supplemental Fig. 7D-H). The similar results were also observed in mice challenged with LLC or B16-OVA cells (Supplemental Fig. 7I-P).

Ogt deficiency inhibits tumor progression through enhancing infiltration by CD8+ T cells.

A-B) Growth volume, weight and survival of Ogt+/+ or Ogt−/− MC38 tumors in C57BL/6J mice, n=5 respectively. C-D) Growth volume, weight and survival of Ogt+/+ or Ogt−/− LLC tumors in C57BL/6J mice, n=5 respectively. E-H) Flow cytometry analysis showing percentage of CD4+ and CD8+ T cells (E), IFN-γ+ (F), TNF-α+ (G), IFN-γ+TNF-α+ double positive (H) expressing intratumoral CD8+ T cells in MC38 tumors, subcutaneous tumor isolated at day 18 post-tumor inoculation, n=5 respectively. I) Growth volume and weight of Ogt+/+ or Ogt−/− MC38 in Rag2-/- mice, n=5 respectively. J-K) Growth volume, weight and survival of Ogt+/+ or Ogt−/− MC38 tumors injected with either control IgG or anti-CD8α at day 0, 7 and 14 post tumor inoculation in C57BL/6J mice, n=5 respectively. L-M) Ogt-/-rescued MC38 tumors in C57BL/6J mice, tumor growth volume and weight (L), survival (M). N-O) Ogt−/−cGAS−/− or Ogt−/−Sting−/− double knockout MC38 tumors in C57BL/6J mice, tumor growth volume and weight (N), survival (O). P) Flow cytometry analysis showing percentage of CD4+ and CD8+ T cells (N), CD8+ IFN-γ+ (O), CD8+ TNF-α+ (P) in Ogt−/−cGAS−/− or Ogt−/−Sting−/− double knockout MC38 tumors in C57BL/6J mice, subcutaneous tumor isolated at day 18 post-tumor inoculation. Q-R) Growth volume, weight and survival of Ogt+/+ or Ogt−/− MC38 tumors injected with either control IgG or anti-PD-L1 at day 7, 10 and 13 post tumor inoculation in C57BL/6J mice, n=5 respectively. S) Kaplan-Meier survival curves for colorectal cancer patients with low (n=207) or high (n=231) OGT transcripts in TCGA dataset. T) Progression-free survival curves for colorectal cancer patients with low (n=58) or high (n=58) OGT transcripts in TCGA dataset. U) Scatterplot presenting the association between the mRNA expression level of OGT and CD8+ T infiltration, Spearman’s r = −0.263, p = 9.75E-6, Spearman’s rank correlation test. Data are representative of two or three independent experiments. Statistical significance was determined by Spearman’s rank correlation test, unpaired Student’s t-test, one-way ANOVA, two-way ANOVA, *p < 0.05, **p < 0.01, ***p < 0.001, ns, no significant difference. Data represent the mean of ±SD.

To determine whether Ogt deficiency is dependent upon adaptive immune system, we inoculated MC38 control and Ogt knockout cells into immunodeficient Rag2−/− mice and tracked tumor growth. The difference of tumor growth rate disappeared between MC38-Ogt knockout and control ones (Fig. 6I), indicating their association with an impaired immune response. We then postulated that Ogt knockout might have potent antitumor effects in vivo through CD4+ T and CD8+ T cells. In order to test this hypothesis, we performed antibody-mediated CD4+ T or CD8+ T cells depletion in Ogt−/− and control tumor-bearing mice and examined the tumor growth and survival. Depletion CD8+ T cells dramatically enhanced the tumor growth in Ogt−/− tumor-bearing mice, compared to the isotype antibody treatment group, and their antitumor activity of Ogt deficiency disappeared, both in tumor volume, weight and survival curve (Fig. 6J-K). However, depletion CD4+ T cells have no obvious effect both in tumor volume and weight in Ogt−/− tumor-bearing mice, compared to the isotype antibody treatment group (Supplemental Fig. 8A). These results implied that CD8+ T cells mediated the inhibitory effect of Ogt deficiency on tumor progression.

Next, to determine whether this effect is specific for OGT knockout, we performed this experiment in vivo using MC38 tumor-bearing rescue model. As we expected, tumor growth and survival phenotype disappeared both in OGT rescued cells (Fig. 6L-M) and in OGT/cGAS or OGT/STING double knockout tumors (Fig. 6N-O). Furthermore, flow cytometry results showed that the proportion of CD4+ T and CD8+ T cells, CD8+ IFN-γ+ and CD8+ TNF-α+ disappeared both in OGT rescued cells (Supplemental Fig. 8B-D) and in OGT/cGAS or OGT/STING double knockout tumors in vivo (Fig. 6P and Supplemental Fig. 8E-F). These results proved that OGT deficiency induces the cGAS/STING and activates the type I IFN pathway.

Increasing evidences support that intratumoral infiltration of CD8+ T cells dictates the response to immune checkpoint blockade (ICB) therapy and its efficacy on various cancers [37, 38]. Blocking PD-L1 can restore the anti-tumor immune function and enhance the antitumor immunity by promoting CD8-positive T-cell infiltration, which is widely used in clinical immunotherapy [39, 40]. Because Ogt deficiency induced tumor cell-intrinsic immune response to recruit CD8+ T cells into MC38, LLC and B16-OVA cells, we hypothesized that Ogt deficiency potentiated enhance the efficacy of PD-L1 blockade in vivo. To test this hypothesis, we carried out the combination treatment of Ogt knockout and neutralizing antibody (anti-PD-L1). MC38 and LLC tumor growth was significantly delayed in tumor-bearing mice treated with PD-L1 antibody compared to isotype control (Fig. 6Q and Supplemental Fig. 8G), which translated into extended survival (Fig. 6R and Supplemental Fig. 8H). Deletion of Ogt synergized with PD-L1 blockade treatments to improve antitumor immunity. We next assessed the potential relevance of OGT in human cancer immunity. We first analyzed gene expression profiles of cancer patients from TCGA database and survival, we found that low OGT expression was associated with improved overall survival (OS) and progression free survival (PFS) in patients with COAD (Fig. 6S-T). Using TIMER2.0 (http://timer.cistrome.org) analysis, we found that OGT expression negatively correlated with CD8+ T cell infiltration (Fig. 6U). For further analysis, we found that a set of genes associated with immune response was robustly regulated in OGT high and OGT low patients. The response interferon-gamma, interferon-gamma production, cellular defense response, regulation of inflammatory response, acute inflammatory response is upregulated; DNA mismatch repair is a downregulated processe, as shown by Gene Ontology (GO) enrichment and pathway analysis (Supplemental Fig. 9A). Gene Set Enrichment Analysis (GSEA) showed that T cell activation, response to interferon-gamma, interferon-gamma production, antigen processing and presentation, interleukin-1/12 production, dectin-1 mediated noncanonical NF-κB signaling are negatively correlated with OGT expression (Supplemental Fig. 9B-H), while mismatch repair, covalent chromatin modification and DNA repair complex are positively correlated with OGT expression (Supplemental Fig. 9I-K). Of our most interest, we found that CD8A, IFNG, ISG15, MX1, CD274 and CXCL10 expression are negatively correlated with OGT expression (Supplemental Fig. 9L-Q). These data suggest a potential involvement of OGT expression and antitumor immunity in patients with cancer.

Combination therapy with OSMI-1 and anti-PD-L1 Ab augmented T cells and antitumor immunity

Based on previous studies, we know that Ogt knockout causes DNA damage, accumulates cytosolic DNA, induces cGAS-STING pathway and activates antitumor immunity. Here we speculated that OGT inhibitor may cause DNA damage and activate antitumor immunity. OSMI-1 is a small molecule inhibitor of OGT that does not significantly affect other glycosyltransferases and is active in a very low doses [41]. The cell proliferation exhibited no obvious difference in different concentrations in MC38 and LLC cells in vitro assay (Supplemental Fig. 10A-B). We stained cytosolic dsDNA with anti-dsDNA and found that treatment with OSMI-1 could significantly induce a high percentage of cytosolic DNA accumulation (Fig. 7A and Supplemental Fig. 10C). We next examined the DNA damage and found that γH2AX was obviously increased in OSMI-1 treated cells (Fig. 7B-C). We also performed the anti-γH2AX immunofluorescence staining and found that OSMI-1 significantly induced immense DNA strand breakage in treated cells (Fig. 7D and Supplemental Fig. 10D).

Combination therapy with OSMI-1 and anti-PD-L1 augmented T cells and antitumor immunity.

A) The anti-dsDNA fluorescence staining treated with 50 μM and 100 μM in MC38 cells, respectively. B-C) Western blot analysis of protein expression treated with 50 μM and 100 μM in MC38 and LLC cells, respectively. D) The anti-γH2AX fluorescence staining treated with 50 μM and 100 μM in MC38 cells, respectively. E-F) Growth volume, weight and survival of MC38 tumors injected with either control OSMI-1 or anti-PD-L1 in C57BL/6J mice. G-H) Growth volume, weight and survival of LLC tumors injected with either control OSMI-1 or anti-PD-L1 in C57BL/6J mice. I-K) Flow cytometry analysis showing percentage of CD4+ and CD8+ T cells (I), CD8+ IFN-γ+ cells (J), CD8+ TNF-α+ cells (K) in MC38 tumors, subcutaneous tumor isolated at day 18 post-tumor inoculation. L-N) Flow cytometry analysis showing percentage of CD4+ and CD8+ T cells (L), CD8+ IFN-γ+ cells (M), CD8+ TNF-α+ cells (N) in LLC tumors, subcutaneous tumor isolated at day 18 post-tumor inoculation. Data are representative of three or four independent experiments. Statistical significance was determined by unpaired Student’s t-test, one-way ANOVA, two-way ANOVA, *p < 0.05, **p < 0.01, ***p < 0.001, ns, no significant difference. Data represent the mean of ±SD.

As we all know, the presence of cytosolic DNA could trigger activation of cGAS-STING pathway [42]. To investigate whether OSMI-1 activated cGAS/STING pathway, we examined activation of major regulators of the pathway in OSMI-1-treated MC38 and LLC cells, as indicated by increased phosphorylation of STAT1, TBK1, and IRF3 and reduced STING expression (Fig.7B-C), which is consistent with Ogt knockout cells. As mentioned earlier, deletion of Ogt synergized with PD-L1 blockade treatments to improve antitumor immunity. We next sought to determine whether OSMI-1 enhanced the antitumor immune effect of anti-PD-L1 antibody in vivo by using MC38 tumor-bearing model. To our surprise, similar to anti-PD-L1 therapy, OSMI-1 alone significantly inhibited MC38 tumor growth and survival, and the combination of OSMI-1 and anti-PD-L1 therapy resulted in superior tumor suppression compared with monotherapy (Fig. 7E-F). Flow cytometry results showed that proportion of CD4+ T and CD8+ T cells was increased both in OSMI-1, anti-PD-L1 treatment alone and combined treatment with OSMI-1 and anti-PD-L1 antibody (Fig. 7I), production of IFN-γ and TNF-α were significantly enhanced in intratumoral CD8+ T cells not only combined treatment with OSMI-1 and anti-PD-L1 antibody, but also single-agent OSMI-1 treatment (Fig. 7J-K). This pharmacological inhibition model is consistent with MC38 Ogt knockout tumor-bearing mice model.

Furthermore, we treated mice bearing LLC tumors, because it represents a most aggressive murine tumor model and are highly resistant to various immunotherapies. Our results showed that treatment with single-agent OSMI-1 can slightly inhibited LLC tumor growth and survival. Combined treatment with OSMI-1 and anti-PD-L1 antibody caused significantly greater tumor suppression than either monotherapy (Fig. 7F-G). We also found that proportion of CD4+ T and CD8+ T cells was significantly increased in combined treatment with OSMI-1 and anti-PD-L1 antibody. However, single-agent OSMI-1 or anti-PD-L1 antibody treatment cannot significantly increased the proportion of CD4+ T and CD8+ T cells (Fig. 7L). Production of IFN-γ and TNF-α were significantly enhanced in intratumoral CD8+ T cells in combined treatment with OSMI-1 and anti-PD-L1 antibody, but not in single-agent OSMI-1 or anti-PD-L1 treatment (Fig. 7M-N). This model is also consistent with Ogt knockout tumor-bearing mice model. In summary, OGT inhibitor OSMI-1 induces DNA damage and cytosolic DNA accumulation, activates cGAS/STING pathway. Combined OGT inhibitor with anti-PD-L1 antibody markedly suppressed tumor growth and increased CD8+ T cells and production of IFN-γ and TNF-α in tumor. These results demonstrated a pivotal role of OGT inhibition in augmenting the antitumor immune response of ICB. Given the increasing importance of immunotherapy for the management of patients with that OGT inhibitors, combined with anti-PD-L1 blockade may offer a particularly attractive strategy for the treatment of colorectal and lung cancer, which are instrumental in turning ‘cold tumors’ into ‘hot tumors’.

Discussion

DNA repair and DNA damage signaling pathways play an important role in the maintenance of genomic stability and integrity through correcting the impaired DNA that may contribute to carcinogenesis [43]. DNA repair pathways are triggered to maintain genomic stability when cells are exposed to endogenous or exogenous DNA-damaging agents. In antitumor therapies, chemotherapy and radiotherapy induce cell death by directly or indirectly causing DNA damage and increase the tumor sensitivity to cancer therapies. DNA damage-inducing therapies tremendously beneficial for cancer therapy and function by the direct or indirect formation of DNA lesions and subsequent inhibition of cellular proliferation. Therefore, targeting DNA repair pathways may be a potential therapeutic approach for cancer treatment. In this study, we found that deletion OGT causes un-controlled expansion of DNA damage and induced cytosolic dsDNA accumulation in tumor cells.

The cytosolic DNA sensing pathway has emerged as the major link between DNA damage and innate immunity, DNA damage in the nucleus results in the accumulation of cytosolic DNA, which activate the cGAS–STING pathway [44–49]. The cGAS–STING pathway is the key cytosolic DNA sensor that is responsible for the type I IFN production, DC activation, and subsequent priming of CD8+ T cells against tumor-associated antigens [30, 50, 51]. Recent evidence shows that proper activation of tumor cell-intrinsic immunity or innate immune cells can enhance antitumor immunity [34, 52, 53]. In this study, Ogt deficiency can activate cGAS-STING pathway and induce the expression of proinflammatory cytokines (e.g., IFNB1 and ISGs) and chemokines (e.g., CXCL10) in a TBK1-IRF3-dependent manner, this cGAS-STING-IIFNB1-CXCLl10 signal axis can provide an activated signal for epitope-specific CD8+ T cell priming by antigen-presenting cells. Finally, it increased the infiltration of tumor CD8+ T lymphocytes and better inhibited tumor growth in TME, while knockout cGAS or STING can eliminate the CD8+ T cells proliferation signal and abolish antitumor immunity (Supplemental Fig. 11).

Immunotherapy with checkpoint-blocking antibodies targeting CTLA-4 and PD-1/PD-L1 have revolutionized cancer treatment and drastically improved the survival of individuals in the clinical treatment. Although immunotherapy make great progress in treatment of solid tumors. However, only around 20% of patients with non-small cell lung cancer (NSCLC) respond to mono-immunotherapy, and a large proportion of individuals develop resistance. Therefore, there is a requirement to explore novel alternative strategies and personalized immunotherapy strategies through a combination of PD-1/PD-L1 with small molecular targets to improve sensitivity to activated antitumor immune response for patients and solve the drug resistance [54, 55]. Here we showed that the OGT inhibitor OSMI-1 was shown to induce DNA damage and cytosolic DNA accumulation which led to activation of the cGAS-STING-TBK1-IRF3 pathway, then enhanced the innate and adaptive immune responses to tumor cells, which reversed the immunosuppressive TME by increasing CD8+ T cells infiltration. To our surprise, especially in lung cancer mice model, combination OSMI-1/PD-L1 therapy can achieve a better antitumor effect than either monotherapy.

In summary, our findings demonstrated deficiency in OGT mediated genomic instability and result in cytosolic dsDNA accumulation, which activating the cGAS-STING signaling pathway, increasing inflammatory cytokines and enhancing antitumor immunity. Our study also addresses an unmet clinical need through the combination of OGT inhibition and anti-PD-L1 therapy, which may represent a promising strategy for colorectal and lung cancer therapy.

eLife Assessment

This valuable study provides the detailed molecular mechanism of how OGT, an O-GlcNac transferase, promotes cancer progression. Using loss-of-function OGT models, the authors demonstrated that OGT cleaves HCF-1, a guardian of genomic stability. These solid findings can lead to some potential approaches to modulate anti-tumor immunity by targeting this process.

Reviewer #1 (Public Review):

Summary:

This study provides the detailed molecular mechanism of how OGT, an O-GlcNac transferase, promotes cancer progression. Using loss-of-function OGT models, the authors demonstrated that OGT cleaves HCF-1, an important guardian of genomic stability. The resulting genomic instability in OGT-knockout tumors leads to cytosolic DNA accumulation, the activation of cGAS-mediated type I IFN responses, and increased CD8+ T cell infiltration into the tumors. Moreover, treatment with OGT inhibitor synergized with anti-PDL1 immune-checkpoint blockade.

Strengths:

Novel findings of how OGT promotes tumor progression.

Weaknesses:

(1) Some of the data is problematic and does not always support the authors' conclusions.

(2) The writing needs significant improvement. In places, it is hard to understand or could mislead the readers.

(3) Figure legends are minimalistic and do not provide sufficient information.

(4) Discussion does not put the findings of this study into a broader context of the field but merely restates them.

Reviewer #2 (Public Review):

Summary:

In this study, the author demonstrates that deficiency or pharmacological inhibition of O-glcNac transferase (OGT) enhances tumor immunity in colorectal cancer models. The authors propose that OGT deficiency triggers a DNA damage response, activating the cGAS-STING innate immunity pathway and promoting a Type I interferon response. They suggest that OGT-mediated processing of HSF1 is crucial in maintaining genomic integrity. This research is significant as it identifies OGT inhibition as a potential immunomodulatory target in cancer treatment.

Strengths:

The strength of the paper lies primarily in the in vivo data, demonstrating the impact of OGT deficiency or inhibition on modulating tumor growth and anti-tumor immunity. The experiments are well-controlled. However, there are several unresolved questions:

Weaknesses:

The mechanisms of how OGT deficiency can trigger DNA damage and the role of this response in promoting immunity are only partially addressed in the manuscript.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Recommendations For The Authors):

Some of the data is problematic and does not always support the authors' conclusions:

(1) Fig. 1K and H are identical.

Thank you for pointing out this problem in manuscript. We apologize for this unintentional mistake and have replaced Fig. 1K.

(2) The graph in Figure 2B contradicts the text. It is not obvious how the image was quantified to produce the histological score graph..

We thank the reviewer for pointing out this problem in manuscript, as the reviewer suggested, we have replaced the Figure 2B.

(3) In Figures 2C and D, there is no clear pattern of changes in pro-inflammatory or anti-inflammatory cytokines, despite the authors' claims in the text.

We appreciate the comment, we think the reason is that the level of cytokines in the tissue is low, so the pattern of changes is not obvious.

(4) It is unclear why the anti-dsDNA antibody does not stain the nucleus in Figure 4B. The staining with anti-dsDNA and DAPI does not match well. Figure 5H shows there is still lots of cytosolic DNA in OGT-/- HCF-1-C, measured by DAPI. These data do not support the authors' conclusion that HCFC600 eliminates cytosolic DNA accumulation (line 229). There is no support for the authors' claim that HCF-1 restrains the cGAS-STING pathway (line 330).

We thank these insightful comments, the most critical step in staining cytosolic DNA is to proceed to a low-permeabilization as to allow the antibody to cross the cellular membrane but not the nuclear membrane, that’s why the anti-dsDNA antibody does not stain the nucleus. In Figure 5H, we think we used a high concentrated DAPI to do the staining and nucleus DNA get stained, looks like it’s the cytosolic DNA.

(5) In Figure 5B, there is no increase in HCF-1 cleavage after OGT over-expression.

We appreciate the reviewer for his/her comment, we think the reason is that we used the cell line to stably overexpress OGT-GFP and we may have missed the time point when the increase in HCF-1 cleavage occurred, so there is no big increase of it. However, there is a significant increase in Figure 5C.

(6) In Figure 7, the TNF-a staining does not inspire confidence.

We thank the reviewer for his/her comment, from both Figure 7K (MC38 tumor model) and Figure 7N (LLC tumor model), we observed a significant increase in TNF-α+ CD8+ T cells in the group treated with the combination of OSMI-1 and anti-PD-L1 compared to the control group, as evidenced by the clear clustering.

The writing needs significant improvement:

(1) There are multiple English grammar mistakes throughout the paper. It is recommended that the authors run the manuscript through an editing service.

We thank the reviewer for his/her suggestion. We apologize for the poor language of our manuscript. We worked on the manuscript for a long time and the repeated addition and removal of sentences and sections obviously led to poor readability. We have now worked on both language and readability and have also involved native English speakers for language corrections. We really hope that the flow and language level have been substantially improved.

(2) Some passages are misleading -- lines 161-162, line 217, lines 241-242, 263-264, 299-300. They need to be changed substantially.

We apologize for these mistakes, we have changed them.

(3) Figure legends should be rewritten. Currently, they are too abbreviated to be understood.

We apologize for that, we have rewritten them.

(4) Discussion should also be thoroughly reworked. Currently, it is merely restating the authors' findings. The authors should put their findings in the broader context of the field.

We apologize for that. For a better understanding of our study, we have reworked the discussion.

Reviewer #2 (Recommendations For The Authors):

(1) Previous studies (DOI: 10.1093/nar/gkw663, 10.1016/j.jgg.2015.07.002, 10.1016/j.dnarep.2022.103394) have suggested that OGT deficiency triggers DNA damage, connecting it to DNA repair and maintenance through various mechanisms. This should be acknowledged in the manuscript. Conversely, the role of HCF1 and its cleaved products in maintaining genomic integrity hasn't been previously shown. The authors investigate HCF1's role solely in the context of OGT inhibition. It is unclear whether this is also true under other stimuli that trigger DNA damage, whether fragments of HCF1 specifically reduce DNA damage, or if HSF1 is involved in the basal machinery that would be defective only in the absence of OGT.

We have acknowledged the manuscript mentioned above. In this paper we focused on the OGT function, which is related to HCF1. The role of HCF1 and its cleaved products in maintaining genomic integrity is an interesting topic, we may focus on it in next project.

(2) In villin-CRE-deficient mice, the authors observe generic inflammation in the intestine unrelated to tumor development. It's unclear if this also occurs in the presence of OGT inhibitors in mice, whether these inhibitors induce a systemic inflammatory (Type I interferon) response, or if certain tissues like the intestine or proliferating tumor cells are more susceptible to such a response.

We thank the comment, yes, investigating whether OGT inhibitors induce an inflammatory response, either systemically or tissue-specifically, is a very interesting project to focus on. However, in our current paper, we use a genetic method to identify the role of OGT deficiency in intestine inflammation-induced tumor development. This approach provides convincing evidence for our hypothesis. We may test the effect of OGT inhibitors on inflammation and tumor development in our next project.

(3) Another critical observation is the magnitude of the interferon response triggered by DNA damage in the OGT-deficient models. While it's known that DNA damage can activate cGAS-STING, the response's extent in the absence of OGT prompts the question of whether additional OGT-specific features could explain this phenomenon. For example, Lamin A, essential for nuclear envelope integrity and shown to be O-glycosylated (DOI: 10.3390/cells7050044), and other components of the nuclear envelope or its repair might be affected by OGT. The impact of OGT inhibition on nuclear envelope integrity compared to other DNA-damaging agents could be explored.

We appreciate the comment, in this project, we find an OGT binding protein, HCF1, though LC–MS/MS assay, it’s a top one candidate in binding profiles, so we focus on it. Like Lamin A and other components of the nuclear envelope still are good targets to check, we may explore these in our next project.

(4) The authors also demonstrate a correlation between OGT expression in tumors compared to healthy tissues. However, the reason is unclear, raising questions about whether this is a consequence of proliferation or metabolic deregulation in the cancer. The authors should address this aspect.

We appreciate the reviewer’s insightful point. It is very good questions and very interesting research. However, in this paper we focused on how OGT influence its downstream molecules to promote tumor, we didn’t check why OGT is increased in tumors, it is not the scope of this current work, we would love to investigate it in the future.

Minor points

Please add the legend to Figures S2, S3 and S5.

We thank the comment, we have added the legend to Figures S2, S3 and S5.

The sentence line 137 should be clarified as OGT deficiency seems more related to increased inflammation in this model.

We thank the comment, we have corrected the sentence line 137.

Line 732 has a ( typo before the number 34.

We thank the comment, we have corrected the sentence line 732.

Version 2

Abstract

Abstract

The O-GlcNAc transferase (OGT) is an essential enzyme that mediates protein O-GlcNAcylation, a unique form of posttranslational modification of many nuclear and cytosolic proteins. Recent studies observed increased OGT and O-GlcNAcylation levels in a broad range of human cancer tissues compared to adjacent normal tissues, indicating a universal effect of OGT in promoting tumorigenesis. Here, we show that OGT is essential for tumor growth in immunocompetent hosts by repressing the cyclic GMP-AMP synthase (cGAS)-dependent DNA sensing pathway. We found that deletion of OGT (Ogt−/−) caused a marked reduction in tumor growth in both syngeneic tumor models and a genetic colorectal cancer (CRC) model induced by mutation of the Apc gene (Apcmin). Pharmacological inhibition or genetic deletion of OGT induced a robust genomic instability (GIN), leading to cGAS-dependent production of the type I interferon (IFN-I) and IFN-stimulated genes (ISGs). As a result, deletion of Cgas or Sting from Ogt−/− cancer cells restored tumor growth, and this correlated with impaired CD8+ T cell-mediated antitumor immunity. Mechanistically, we found that OGT-dependent cleavage of host cell factor C1 (HCF-1) is required for the avoidance of GIN and IFN-I production in tumors. In summary, our results identify OGT-mediated genomic stability and activate cGAS-STING pathway as an important tumor cell-intrinsic mechanism to repress antitumor immunity.

Introduction

Cancer cells can maintain malignant phenotypes partially due to altering the post-translational modification (PTM) patterns of cancer-related functional proteins under the stimulation of extracellular and intracellular factors [1]. Protein modification by the O-linked β-N-acetylglucosamine (O-GlcNAc) is a dynamic and reversible post-translational modification, which is added to the hydroxyl group of a specific serine or threonine residue in a target protein by O-GlcNAc transferase (OGT) and removed by O-GlcNAcase (OGA) [2]. O-GlcNAcylation is a fast-cycle and nutrient-sensitive PTM, which modifies thousands of cytoplasmic, nuclear, and mitochondrial proteins and mediates crosstalk with protein phosphorylation, regulating signal transduction and affecting protein localization, activity, stability, and protein-protein interaction. Dysregulation of O-GlcNAcylation is associated with multiple metabolic diseases and cancer [3, 4].

Recent studies observed increased OGT and O-GlcNAcylation level in human colon cancer tissues compared to adjacent normal tissues [5–7], indicating an essential role of OGT-mediated protein O-GlcNAcylation in the pathogenesis of colon cancer. Several oncogenic proteins that are involved in the pathogenesis of colon cancer have been shown to be directly modified by O-GlcNAc, including β-catenin and NF-κB [8–10]. For example, O-GlcNAcylation of β-catenin at T41 inhibits its phosphorylation, which subsequently attenuates its ubiquitination and degradation and promotes oncogenic activity [7, 11]. In contrast, inhibitory roles of O-GlcNAc signaling in the growth of human colon cancer and in oncogenic Wnt/β-catenin signaling have also been reported [12, 13]. The aforementioned studies show that O-GlcNAc modification of specific proteins can play opposing roles in tumorigenesis. However, the overall effect of OGT-mediated O-GlcNAcylation in cancer remains unknown.

The cGAS/STING cytosolic DNA-sensing pathway plays a vital role in activating the innate immune response and production of the type I interferons (IFN-I) [14]. Cyclic guanosine monophosphate (GMP)–adenosine monophosphate (AMP) synthase (cGAS) interacts with cytosolic double-stranded DNA (dsDNA) in a sequence-independent manner. The direct binding of cGAS to cytosolic dsDNA promotes cGAS homodimerization and activates the catalytic activity of cGAS, producing 2’, 3’-cyclic GMP-AMP (cGAMP) from ATP and GTP. The second messenger cGAMP binds to and activates the endoplasmic resident stimulator of interferon genes (STING). Once activated, STING translocates to the ER-Golgi intermediate compartment (ERGIC) to recruit TANK-binding kinase 1 (TBK1) and IFN regulatory factor 3 (IRF3), leading to the production of IFN-I and activation of numerous IFN-stimulated genes (ISGs) [15–17]. However, it remains largely unknown whether OGT expression affects cGAS-STING pathway and antitumor immunity.

In this study, we find that deficiency or pharmacological inhibition of OGT and subsequent accumulation of cytosolic dsDNA activates the cGAS-STING pathway and induces CD8+ T cell-dependent antitumor immunity. Deletion of cGAS or STING diminishes DNA sensing and lead to progressive tumor growth. Mechanistically, we show that OGT could interacts with HCF-1 and cleaves it, which contributes to the maintenance of genomic stability. Re-expression of HCF-1C600 in Ogt−/− tumor cells inhibit production of cytosolic dsDNA and IFN-I. In summary, our findings demonstrate that OGT-mediated DNA damage and activate cGAS-STING pathway as an important tumor cell-intrinsic mechanism to repress antitumor immunity and provides a window for potential therapeutic opportunities for in OGT-dependent cancer.

Results

Increased OGT expression in human and mouse tumor samples

Protein O-GlcNAcylation is upregulated in various cancers [12, 18–21]. OGT is the only known enzyme that mediates O-GlcNAcylation of proteins at the Ser or Thr residues [22], we hypothesized that OGT could serve as an important regulator to regulate cancer cell growth and serve as a biomarker for cancer. Initial analysis of data from The Cancer Genome Atlas (TCGA) dataset in GEPIA2 (http://gepia2.cancer-pku.cn/#index) and found a significant positive correlation between OGT mRNA expression and tumorigenesis in bladder urothelial carcinoma (BLCA), cholangiocarcinoma (CHOL), colon adenocarcinoma (COAD), esophageal carcinoma (ESCA), head and neck squamous cell carcinoma (HNSC), kidney chromophobe (LIHC), lung adenocarcinoma (LUAD), prostate adenocarcinoma (PRAD), rectum adenocarcinoma (READ), sarcoma (SARC), and a stomach adenocarcinoma (STAD) (Supplemental Fig. 1A). Next, we utilized UALCAN database analysis, and found that OGT mRNA expression was compared between 286 COAD samples and 41 adjacent or normal samples, the expression of OGT was significantly increased in COAD at the transcriptional level (Fig. 1A). A significant positive correlation tendency between OGT mRNA expression and individual different tumor stages was observed. The stage IV COAD tissues exhibited the highest expression level of OGT in compared with low stage (Fig. 1B). We also observed a significant positive correlation between OGT mRNA expression and nodal metastasis status (Fig. 1C). Furthermore, we found that the protein level of OGT was also significantly increased in COAD patient samples and amongst individual different stages based on the CPTAC and HPA online database (Fig. 1D-F). Similar results are found in LUAD (Supplemental Fig. 1B-F). These results suggest that high mRNA and protein levels of OGT in tumorigenesis were consistent in different databases.

OGT is significantly upregulated in human and mouse tumor samples.

(A-C) Boxplot showing mRNA expression level of Ogt, Normal and tumor samples (A), Individual stages (B), Nodal metastasis status (C). The plot was generated using the UALCAN online server. D-E) Boxplot showing protein expression level of OGT, Normal and tumor samples (A), Individual stages (B). The plot was generated using the UALCAN online server (https://ualcan.path.uab.edu/analysis.html). F) IHC analysis of OGT expression in normal colon tissues, primary colon tumor samples (from Human Protein Atlas, https://www.proteinatlas.org/), scale bar: 400μm. G) Western blot analysis of OGT and O-GlcNAc expression in normal, adjacent and tumor tissues in Apcmin spontaneous tumor mice. H) HE and IHC staining of OGT in adjacent and tumor tissues in Apcmin spontaneous tumor mice, scale bar: left panel 275 μm, right panel 75μm, n=3 respectively. I) Schematic of AOM/DSS model of colitis-associated colorectal cancer (CAC). J) Western blot analysis of OGT and O-GlcNAc expression in normal, adjacent and tumor tissues in CAC model. K) HE and IHC staining of OGT in adjacent and tumor tissues in CAC model, scale bar: left panel 275 μm, right panel 75μm, n=4 respectively. human samples (A-F), mouse samples (G-K). Statistical significance was determined by Pearson test, unpaired Student’s t-test, *p < 0.05, **p < 0.01, ***p < 0.001, ns, no significant difference. Data represent the mean of ± SD.

To determine whether intestinal OGT expression was increased in Apcmin colorectal tumor mouse model, intestinal tissues were collected for western blot and immunohistochemical staining (IHC) analysis. Intestinal OGT protein was markedly increased in Apcmin mouse tumor tissues compared to adjacent or normal tissues (Fig. 1G). As expected, O-GlcNAcylation proteins levels in intestinal tissues were also significantly higher in tumor tissues than adjacent or normal tissues. IHC staining revealed that mouse OGT protein was markedly higher in tumor tissues than in adjacent tissues (Fig. 1H), which was consistent with the expression pattern in human samples. OGT protein was also upregulated in the azoxymethane (AOM)/dextran sodium sulfate (DSS) colorectal tumor model (Fig. 1I-K). Together, these data strongly suggest that OGT may play a critical role in tumorigenesis and serve as a prognostic marker and therapeutic target in cancer treatment.

Epithelial OGT deletion inhibits mouse colorectal tumorigenesis

To determine whether OGT may be an important therapeutic target for tumor treatment, we generated an intestinal epithelial cell-specific Ogt deletion (OgtΔIEC) by crossing Ogtfl/fl mice with Villin-Cre mice. OgtΔIEC mice were then crossed with Apcmin mice to generate OgtΔIECApcmin mice (Supplemental Fig. 2). To determine whether intestinal OGT expression was decreased in OgtΔIECApcmin mice, intestinal tissues were collected for western blot analysis. Intestinal OGT levels were drastically decreased in OgtΔIECApcmin mice compared to wildtype Apcmin mice (Fig. 2A). Importantly, deletion of the Ogt in intestinal epithelial cells resulted in significantly reduced tumor size and total number of polyps at 20 weeks of age (Fig. 2A), indicating that OGT promotes oncogenic transformation in colorectal tumor in vivo. To gain insight into the role of protein on intestinal carcinogenesis in intestinal cancer, we examined the differences in intestinal carcinogenesis. Hematoxylin and eosin (H&E) staining demonstrated that epithelial inflammation increased in OgtΔIECApcmin mice compared to wildtype Apcmin mice (Fig. 2B). OGT deficiency in intestinal tissues was also associated with significantly elevated gene expression of key pro-inflammatory cytokines like interleukin Il1α, Il6, TNF-α, as well as several interferons and ISGs like Isg15, Mx1, Cxcl10 (Fig. 2C). Furthermore, we found IL-2, IL-6, IL-10, IL-17, IFN-α, IFN-β, IFN-γ and CXCL10 were upregulated compared with control mice (Fig. 2D).

Epithelial OGT deletion inhibits mouse colorectal tumorigenesis.

A) Western blot analysis of OGT expression in intestinal tissues and counting tumor numbers in APCmin and Ogt△IEC mice. B) Histology analysis of intestinal carcinogenesis by HE staining, scale bar: up panel 275 μm, bottom panel 20 μm, n=3 respectively. C) Real-time PCR analysis of cytokines mRNA expression in intestine. D) ELISA analysis of cytokines expression in intestine. Statistical significance was determined by unpaired Student’s t-test, *p < 0.05, **p < 0.01, ***p < 0.001, ns, no significant difference. Data represent the mean of ± SD.

OGT deficiency activates cGAS/STING-dependent IFN-I pathway

Based on previous studies, we found that several interferons and several interferon-stimulated genes, such as Isg15, Mx1, IFN-α, IFN-β and Cxcl10 were upregulated. We hypothesized that OGT deficiency may activate the type I IFN pathway. First, we generated Ogt knock outs of murine colorectal carcinoma models like MC38, LLC, B16-OVA and of a human colorectal cell line, HT29. We then evaluated the expression changes of Ifna4, Ifnb1, Isg15, Mx1 and found that all the genes except Ifna4 mRNA expression were significantly increased in all cultured Ogt knockout cells (Fig. 3A-D). Next, we investigated how the type I IFN pathway in Ogt knockout cells was activated [23, 24]. We found that while the phosphorylation of STAT1, TBK1 and IRF3 is increased, the STING expression is reduced in Ogt knockout MC38, LLC, B16-OVA and HT29 cells (Fig. 3E-F). To determine whether this effect is specific for Ogt knockout, we stably expressed exogenous OGT in Ogt knockout cells and found that the type I IFN pathway activation effect is abolished (Fig. 3F). The type I IFN pathway is typically activated by both viral RNA and dsDNA. In order to eliminate the type I IFN pathway activate independent on RNA but dependent on dsDNA, we knocked out MAVS protein in Ogt−/− cells, which is a RNA pivotal adaptor protein activate the downstream protein kinase TBK1, IFN pathway and ISGs in Ogt−/− cells (24), and we found that knockout of MAVS in Ogt−/− cells has no effect on either Isg15, Mx1, Ifna4 and Ifnb1 expression or the type I IFN pathway (Fig. 3G-H).

OGT deficiency induces cGAS/STING-dependent the type I IFN pathway.

A-D) Real-time PCR analysis of cytokines mRNA expression in different Ogt−/− cell lines including MC38 (A), LLC (B), HT29 (C), B16-OVA (D) cells. E) Western blot analysis of the activation of the interferon signaling pathway in different Ogt−/− cell lines including MC38, LLC and B16-OVA cells. F) Western blot analysis of the activation of the interferon signaling pathway in Ogt−/− rescued cell lines including MC38, LLC, HT29 and B16-OVA cells. G-H) Real-time PCR and western blot analysis of cytokines mRNA expression and the activation of the interferon signaling pathway in different Ogt−/−Mavs−/− double knockout clones in MC38 cells. I-K) Real-time PCR and ELISA analysis of cytokines mRNA expression in Ogt−/−cGAS−/− double knockout clones in MC38 (I-J), HT29 (K) cells. L) Western blot analysis of the activation of the interferon signaling pathway in Ogt−/−cGAS−/− or Ogt−/−Sting−/− double knockout clones in MC38, HT29 and B16-OVA cells. M-N) BMDCs pre-treated with B16-OVA-Ogt−/− (L), B16-OVA-Ogt−/−cGAS−/− or B16-OVA-Ogt−/−Sting−/− cells (M) supernatant, and co-cultured with OT-1 T cell, then T cell proliferation was evaluated by flow cytometry, OVA257-264 as a positive control. Representative fluorescence-activated cell sorting histograms and statistical data are shown. Data are representative of two or three independent experiments. Statistical significance was determined by unpaired Student’s t-test, one-way ANOVA, two-way ANOVA, *p < 0.05,**p < 0.01,***p < 0.001, ns, no significant difference. Data represent the mean of ± SD.

We further investigated the mechanism of the type I IFN signaling activation and found that Ogt knockout induced the reduction of the STING. This result is similar to previous studies [25, 26]. The phenotype of Ogt knockout implies that STING is involved in the activation of the type I IFN pathway. We hypothesized that OGT negatively regulates type I IFN through the cGAS/STING pathway. Next, we produced the OGT/cGAS and OGT/STING double knockout cells both in MC38, HT29 and B16-OVA. Surprisingly, these type I IFN and Cxcl10 signals disappeared in OGT/cGAS and OGT/STING double knockout cells both in mRNA and protein level (Fig. 3I-L).

The type I IFN signaling is a key pathway that promotes antigen presentation and DC activation [27–29]. Phagocytosis of extracellular tumor DNA by DCs triggers the activation of the cGAS-STING-IFN pathway [30, 31]. To determine whether the increased cGAS-STING-IFN signaling in Ogt knockout cells provides an activated signal for single epitope-specific CD8+ T cell priming by antigen-presenting cells (APCs), we added the supernatant from B16-OVA tumor cells into the co-culture system of BMDCs and OT-I cells. The results showed that BMDCs pre-cultured with Ogt+/+ or Ogt−/− B16-OVA tumor cells supernatant provided a potent activated signal for optimal single epitope-specific T cell proliferation (Fig. 3M), while cGAS or STING deficiency in Ogt−/− tumor cells diminished such effects (Fig. 3N). Further, we added the supernatant from Ogt+/+ or Ogt−/− B16-OVA tumor cells into the co-culture system of Ifnar1 knockout BMDCs and OT-I cells. The cell proliferation was abolished when treated with Ifnar1-/-BMDCs pre-cultured with Ogt+/+ or Ogt−/− B16-OVA tumor cells supernatant. (Supplemental Fig. 3). These results demonstrated that Ogt deficiency activate antitumor CD8+ T cells response is dependent on the type I IFN signal in dendritic cells and in a manner dependent on the tumor-cell-intrinsic cGAS-STING pathway.

OGT deficiency causes DNA damage and cytosolic DNA accumulation

Because the accumulation of cytosolic DNA is a consequence of nuclear DNA damage, which can activate immune response. The cGAS senses cytoplasmic DNA as a consequence of nuclear DNA damage [32]. Previous studies showed that the STING is degraded while cGAS-STING is activated [25]. As mentioned above, the STING degraded in Ogt knockout cells. We hypothesized that whether the cytosolic DNA is accumulated in Ogt knockout cells, we stained cytosolic double-strand DNA (dsDNA) with PicoGreen, a widely used immunofluorescence staining that selectively binds to dsDNA [33, 34], and found that a significantly higher percentage of Ogt deficiency cells than control cells in two different Ogt knockout MC38 and LLC cell clones, respectively (Fig. 4A and Supplemental Fig. 4A). In addition, we also stained dsDNA with anti-dsDNA immunofluorescence and the results were similar to PicoGreen staining (Fig. 4B and Supplemental Fig. 4B). We next assessed the phosphorylation of H2AX at Ser 139 (γH2AX), an indirect marker of DNA DSBs in the cell lines [35]. Comparable expression levels of H2AX, we found that γH2AX was dramatically increased in Ogt knockout MC38, LLC, HT29 and B16-OVA cells (Fig. 4C and 4F). We also used the anti-γH2AX immunofluorescence staining and comet assays, a classical quantifying and analyzing DNA damage, and found that Ogt knockout significantly induced immense DNA strand breakage in MC38 cells (Fig. 4D-E). Finally, the rescued results showed that exogenous OGT expression abolished the γH2AX expression both in MC38, LLC, HT29 and B16-OVA cells and DNA damage in MC38 cells (Fig. 4F-H). Together, these data indicated that Ogt knockout caused DNA damage and induced the cytosolic DNA accumulation.

OGT deficiency causes DNA damage and accumulates cytosolic DNA.

A) The extranuclear dsDNA in different Ogt−/− MC38 cells clones were determined by PicoGreen staining assay was quantified by image J. B) The extranuclear dsDNA in different Ogt−/− MC38 cells clones were determined by anti-dsDNA fluorescence staining assay and was quantified by image J. C) Western blot analysis of γH2AX and H2AX expression in different Ogt−/− cell lines including MC38, LLC and B16-OVA cells. D) Analysis of γH2AX and H2AX expression in different Ogt−/− clones by anti-γH2AX staining assay and was quantified by image J. E) The DNA damage was determined by comet assay, and extranuclear dsDNA was analyzed by using CometScore in Ogt−/− MC38 cells. F) Western blot analysis of γH2AX and H2AX expression in Ogt−/− rescued cells including MC38, LLC, HT29 and B16-OVA cells. G) The DNA damage in rescued MC38 cells were determined by comet assay, and extranuclear dsDNA was analyzed by using CometScore. Data are representative of three or four independent experiments. Statistical significance was determined by unpaired Student’s t-test, one-way ANOVA, *p < 0.05, **p < 0.01, ***p < 0.001, ns, no significant difference. Data represent the mean of ± SD.

The C-terminus of HCF-1 rescues DNA damage and IFN-I pathway in Ogt−/− cells

Previous studies in Fig. 3F and Fig. 4F showed that expression of exogenous OGT abolished the type I IFN and γH2AX signals. To explore the mechanism how OGT regulate the IFN and γH2AX production. Next, we use GFP agarose immunoprecipitation and liquid chromatography coupled to tandem MS (LC-MS/MS) to identify different proteins that interacted with OGT in human colonic cancer cell line HT29. Interestingly, HCF-1 was the most enriched protein in the precipitates from OGT restored HT29 knockout cells compared to OGT knockout cells reconstituted with empty vector, based on the number of peptides (indicating the identification confidence) and the number of peptide-spectrum matches (PSMs, indicating the abundance) (Fig. 5A and Supplemental Table 4). Previous studies showed that the human epigenetic cell-cycle regulator (HCF-1) undergoes an unusual proteolytic maturation by OGT cleavage and process resulting in stably associated HCF-1N1011 and HCF-1C600 subunits that regulate different aspects of the cell cycle [36]. We found similar results in this study by using co-immunoprecipitation assay (Fig. 5B-C). We rescued the HCF-1 cleavage phenotype by expression of exogenous OGT (Fig. 5D). To clarify how to OGT regulate function of HCF-1, we transfected empty vector (EV), the full-length HCF-1-HA (HCF-1FL), HCF-11-1011-HA (HCF-1N1011), HCF-11-450-HA (HCF-1N450), HCF-1450-1011-HA (HCF-1N450-1011) and HCF-11436-2035-HA (HCF-1C600) and myc-OGT and found that OGT can bind HCF-1C600 (Fig. 5E), furthermore OGT directly bind the HCF-1C600 (Supplemental Fig. 5A-B). To further understand the physiological function of HCF-1, we transfected EV, HCF-1FL, HCF-1N1011and HCF-1C600 into MC38 control and OGT knockout cells, respectively. As shown in Fig. 5F, transfected HCF-1FL and HCF-1N1011 didn’t rescue the gene expression differences, such as Isg15, Mx1 and Ifnb1, but HCF-1C600 can restore the gene expression differences. These results were confirmed by using western blot and anti-dsDNA immunofluorescence staining (Fig. 5G-H). Overall, these data indicated that OGT regulates the HCF-1 cleavage and maturation, HCF-1C600 can eliminate the cytosolic DNA accumulation, DNA damage, the type I IFN activation and restrain cGAS-STING-mediated DNA sensing.

The C terminal of HCF-1 rescue DNA damage and the type I IFN pathway in Ogt−/− cells.

A) Volcano plot of OGT binding proteins identified by LC–MS/MS from stably expressed exogenous GFP-OGT in OGT knockout HT29 cells. B) OGT and HCF1 binding was confirmed by immunoprecipitation assay in OGT rescued HT29 cells. C) OGT and HCF1 binding was confirmed by immunoprecipitation assay in 293T cells. D) HCF1 cleavage was confirmed by western blot in Ogt rescued MC38 cells. E) Co-IP analysis of the interaction between OGT and different HCF-1 mutant. F) Real-time PCR analysis of cytokines mRNA expression effected by HCF-1 isoforms in MC38 OGT knockout cells. G) Western blot analysis of γH2AX and H2AX expression in exogenous HCF-1C600 expressed MC38 Ogt knockout cells. H) The extranuclear dsDNA were determined by anti-dsDNA fluorescence staining assay and was quantified by image J in exogenous HCF-1C600 expressed MC38 OGT knockout cells. Data are representative of three or four independent experiments. Statistical significance was determined by one-way ANOVA, two-way ANOVA, *p < 0.05, **p < 0.01, ***p < 0.001, ns, no significant difference. Data represent the mean of ± SD.

OGT deficiency inhibits tumor progression through enhancing infiltration by CD8+ T cells

We examined whether inhibiting Ogt can delay the tumor growth and prolong the survival. To further verify these effects, we utilized transplanted tumor model. We found that the Ogt deficiency MC38 colorectal tumor cells have no obvious inhibit cell growth in vitro (Supplemental Fig. 6A). However, we found that a significant delay in tumor growth, tumor weight and prolong mice survival compared to the control group (Fig. 6A-B). We also used lewis lung carcinoma (LLC) cells and the B16-OVA melanoma cells because they represent an aggressive murine tumor model and are highly resistant to various immunotherapies, similar results were shown in Fig. 6C-D and Supplemental Fig. 6B-E. These results showed that Ogt deficiency could delay tumor growth and prolong survival of mice in MC38, LLC, and B16-OVA tumor model. We next examined whether the enhanced antitumor function is related to tumor microenvironment (TME). Consistently, tumors from C57BL/6 immunocompetent mice bearing Ogt−/− MC38 tumors compared with MC38 control tumors, showed higher proportion of CD8+ and CD4+ T cells, functional CD8+ IFN-γ+, CD8+ TNF-α+ and CD8+ IFN-γ+ TNF-α+ double positive T cells, but not CD45+, CD11b+ CD11c+, CD11b+ F4/80+, CD11b+ Ly6C+ and Treg cells (Fig. 6E-H and Supplemental Fig. 7D-H). The similar results were also observed in mice challenged with LLC or B16-OVA cells (Supplemental Fig. 7I-P).

Ogt deficiency inhibits tumor progression through enhancing infiltration by CD8+ T cells.

A-B) Tumor volume, weight of Ogt+/+ or Ogt−/− MC38 tumors in C57BL/6J mice and mice survival, n=5 respectively. C-D) Tumor volume, weight of Ogt+/+ or Ogt−/− LLC tumors in C57BL/6J mice and mice survival, n=5 respectively. E-H) Flow cytometry analysis of percentage of CD4+ and CD8+ T cells population(E) and IFN-γ+ (F), TNF-α+ (G), IFN-γ+TNF-α+ double positive (H) intratumoral CD8+ T cells population in MC38 tumors, subcutaneous tumor isolated at day 18 post-tumor inoculation, n=5 respectively. I) Tumor volume and weight of Ogt+/+ or Ogt−/− MC38 in Rag2-/- mice, n=5 respectively. J-K) Tumor volume, weight of Ogt+/+ or Ogt−/− MC38 tumors injected with either control IgG or anti-CD8α at day 0, 7 and 14 post tumor inoculation in C57BL/6J mice and mice survival, n=5 respectively. L-M) Tumor volume, weight of Ogt-/- rescued MC38 tumors in C57BL/6J mice, tumor growth volume and weight (L), mice survival (M). N-O) Tumor volume, weight of Ogt−/−cGAS−/− or Ogt−/−Sting−/− double knockout MC38 tumors in C57BL/6J mice, tumor growth volume and weight (N), mice survival (O). P) Flow cytometry analysis showing percentage of CD4+ and CD8+ T cells population (N), CD8+ IFN-γ+ (O), CD8+ TNF-α+ T cell population (P) in Ogt−/−cGAS−/− or Ogt−/−Sting−/− double knockout MC38 tumors in C57BL/6J mice, subcutaneous tumor isolated at day 18 post-tumor inoculation. Q-R) Tumor volume, weight of Ogt+/+ or Ogt−/− MC38 tumors injected with either control IgG or anti-PD-L1 at day 7, 10 and 13 post tumor inoculation in C57BL/6J mice and mice survival, n=5 respectively. S) Kaplan-Meier survival curves for colorectal cancer patients with low (n=207) or high (n=231) OGT transcripts in TCGA dataset. T) Progression-free survival curves for colorectal cancer patients with low (n=58) or high (n=58) OGT transcripts in TCGA dataset. U) Scatterplot presenting the association between the mRNA expression level of OGT and CD8+ T infiltration, Spearman’s r = −0.263, p = 9.75E-6, Spearman’s rank correlation test. Data are representative of two or three independent experiments. Statistical significance was determined by Spearman’s rank correlation test, unpaired Student’s t-test, one-way ANOVA, two-way ANOVA, *p < 0.05, **p < 0.01, ***p < 0.001, ns, no significant difference. Data represent the mean of ± SD.

To determine whether Ogt deficiency is dependent upon adaptive immune system, we inoculated MC38 control and Ogt knockout cells into immunodeficient Rag2−/− mice and tracked tumor growth. The difference of tumor growth rate disappeared between MC38-Ogt knockout and control ones (Fig. 6I), indicating their association with an impaired immune response. We then postulated that Ogt knockout might have potent antitumor effects in vivo through CD4+ T and CD8+ T cells. In order to test this hypothesis, we performed antibody-mediated CD4+ T or CD8+ T cells depletion in Ogt−/− and control tumor-bearing mice and examined the tumor growth and survival. Depletion CD8+ T cells dramatically enhanced the tumor growth in Ogt−/− tumor-bearing mice, compared to the isotype antibody treatment group, and their antitumor activity of Ogt deficiency disappeared, both in tumor volume, weight and survival curve (Fig. 6J-K). However, depletion CD4+ T cells have no obvious effect both in tumor volume and weight in Ogt−/− tumor-bearing mice, compared to the isotype antibody treatment group (Supplemental Fig. 8A). These results implied that CD8+ T cells mediated the inhibitory effect of Ogt deficiency on tumor progression.

Next, to determine whether this effect is specific for OGT knockout, we performed this experiment in vivo using MC38 tumor-bearing rescue model. As we expected, decreased tumor growth and prolong mice survival phenotype disappeared both in OGT rescued cells (Fig. 6L-M) and in OGT/cGAS or OGT/STING double knockout tumors (Fig. 6N-O). Furthermore, flow cytometry results showed that the proportion of CD4+ T and CD8+ T cells, CD8+ IFN-γ+ and CD8+ TNF-α+ disappeared both in OGT rescued cells (Supplemental Fig. 8B-D) and in OGT/cGAS or OGT/STING double knockout tumors in vivo (Fig. 6P and Supplemental Fig. 8E-F). These results proved that OGT deficiency induces the cGAS/STING and activates the type I IFN pathway.

Increasing evidences support that intratumoral infiltration of CD8+ T cells dictates the response to immune checkpoint blockade (ICB) therapy and its efficacy on various cancers [37, 38]. Blocking PD-L1 can restore the anti-tumor immune function and enhance the antitumor immunity by promoting CD8-positive T-cell infiltration, which is widely used in clinical immunotherapy [39, 40]. Because Ogt deficiency induced tumor cell-intrinsic immune response to recruit CD8+ T cells into MC38, LLC and B16-OVA cells, we hypothesized that Ogt deficiency potentiated enhance the efficacy of PD-L1 blockade in vivo. To test this hypothesis, we carried out the combination treatment of Ogt knockout and neutralizing antibody (anti-PD-L1). MC38 and LLC tumor growth was significantly delayed in tumor-bearing mice treated with PD-L1 antibody compared to isotype control (Fig. 6Q and Supplemental Fig. 8G), which translated into extended survival (Fig. 6R and Supplemental Fig. 8H). Deletion of Ogt synergized with PD-L1 blockade treatments to improve antitumor immunity. We next assessed the potential relevance of OGT in human cancer immunity. We first analyzed gene expression profiles of cancer patients from TCGA database and survival, we found that low OGT expression was associated with improved overall survival (OS) and progression free survival (PFS) in patients with COAD (Fig. 6S-T). Using TIMER2.0 (http://timer.cistrome.org) analysis, we found that OGT expression negatively correlated with CD8+ T cell infiltration (Fig. 6U). For further analysis, we found that a set of genes associated with immune response was robustly regulated in OGT high and OGT low patients. The response interferon-gamma, interferon-gamma production, cellular defense response, regulation of inflammatory response, acute inflammatory response is upregulated; DNA mismatch repair is a downregulated processe, as shown by Gene Ontology (GO) enrichment and pathway analysis (Supplemental Fig. 9A). Gene Set Enrichment Analysis (GSEA) showed that T cell activation, response to interferon-gamma, interferon-gamma production, antigen processing and presentation, interleukin-1/12 production, dectin-1 mediated noncanonical NF-κB signaling are negatively correlated with OGT expression (Supplemental Fig. 9B-H), while mismatch repair, covalent chromatin modification and DNA repair complex are positively correlated with OGT expression (Supplemental Fig. 9I-K). Of our most interest, we found that CD8A, IFNG, ISG15, MX1, CD274 and CXCL10 expression are negatively correlated with OGT expression (Supplemental Fig. 9L-Q). These data suggest a potential involvement of OGT deficiency and antitumor immunity in patients with cancer.

Combination therapy with OSMI-1 and anti-PD-L1 Ab augmented T cells and antitumor immunity

Based on previous studies, we know that Ogt knockout causes DNA damage, accumulates cytosolic DNA, induces cGAS-STING pathway and activates antitumor immunity. Here we speculated that OGT inhibitor may cause DNA damage and activate antitumor immunity. OSMI-1 is a small molecule inhibitor of OGT that does not significantly affect other glycosyltransferases and is active in a very low doses [41]. The cell proliferation exhibited no obvious difference in different concentrations in MC38 and LLC cells in vitro assay (Supplemental Fig. 10A-B). We stained cytosolic dsDNA with anti-dsDNA and found that treatment with OSMI-1 could significantly induce a high percentage of cytosolic DNA accumulation (Fig. 7A and Supplemental Fig. 10C). We next examined the DNA damage and found that γH2AX was obviously increased in OSMI-1 treated cells (Fig. 7B-C). We also performed the anti-γH2AX immunofluorescence staining and found that OSMI-1 significantly induced immense DNA strand breakage in treated cells (Fig. 7D and Supplemental Fig. 10D).

Combination therapy with OSMI-1 and anti-PD-L1 augmented T cells and antitumor immunity.

A) The extranuclear dsDNA were determined by anti-dsDNA fluorescence staining treated with 50 μM and 100 μM OSMI in MC38 cells respectively and was quantified by image J. B-C) Western blot analysis of protein expression in MC38 and LLC cells treated with 50 μM and 100 μM OSMI, respectively. D) Analysis of γH2AX and H2AX expression by anti-γH2AX staining treated with 50 μM and 100 μM OSMI in MC38 cells and was quantified by image J. E-F) Tumor volume, weight of MC38 tumors injected with either control OSMI-1 or anti-PD-L1 in C57BL/6J mice and mice survival. G-H) Tumor volume, weight of LLC tumors injected with either control OSMI-1 or anti-PD-L1 in C57BL/6J mice and mie survival. I-K) Flow cytometry analysis showing percentage of CD4+ and CD8+ T cells population (I), and CD8+ IFN-γ+ cells (J), CD8+ TNF-α+ cells (K) population in MC38 subcutaneous tumor isolated at day 18 post-tumor inoculation. L-N) Flow cytometry analysis showing percentage of CD4+ and CD8+ T cells population (L), and CD8+ IFN-γ+ cells (M), CD8+ TNF-α+ cells (N) population in LLC subcutaneous tumor isolated at day 18 post-tumor inoculation. Data are representative of three or four independent experiments. Statistical significance was determined by unpaired Student’s t-test, one-way ANOVA, two-way ANOVA, *p < 0.05, **p < 0.01, ***p < 0.001, ns, no significant difference. Data represent the mean of ± SD.

As we all know, the presence of cytosolic DNA could trigger activation of cGAS-STING pathway [42]. To investigate whether OSMI-1 activated cGAS/STING pathway, we examined activation of major regulators of the pathway in OSMI-1-treated MC38 and LLC cells, as indicated by increased phosphorylation of STAT1, TBK1, and IRF3 and reduced STING expression (Fig.7B-C), which is consistent with Ogt knockout cells. As mentioned earlier, deletion of Ogt synergized with PD-L1 blockade treatments to improve antitumor immunity. We next sought to determine whether OSMI-1 enhanced the antitumor immune effect of anti-PD-L1 antibody in vivo by using MC38 tumor-bearing model. To our surprise, similar to anti-PD-L1 therapy, OSMI-1 alone significantly inhibited MC38 tumor growth and survival, and the combination of OSMI-1 and anti-PD-L1 therapy resulted in superior tumor suppression compared with monotherapy (Fig. 7E-F). Flow cytometry results showed that proportion of CD4+ T and CD8+ T cells was increased both in OSMI-1, anti-PD-L1 treatment alone and combined treatment with OSMI-1 and anti-PD-L1 antibody (Fig. 7I), production of IFN-γ and TNF-α were significantly enhanced in intratumoral CD8+ T cells not only combined treatment with OSMI-1 and anti-PD-L1 antibody, but also single-agent OSMI-1 treatment (Fig. 7J-K). This pharmacological inhibition model is consistent with MC38 Ogt knockout tumor-bearing mice model.

Furthermore, we treated mice bearing LLC tumors, because it represents a most aggressive murine tumor model and are highly resistant to various immunotherapies. Our results showed that treatment with single-agent OSMI-1 can slightly inhibited LLC tumor growth and survival. Combined treatment with OSMI-1 and anti-PD-L1 antibody caused significantly greater tumor suppression than either monotherapy (Fig. 7F-G). We also found that proportion of CD4+ T and CD8+ T cells was significantly increased in combined treatment with OSMI-1 and anti-PD-L1 antibody. However, single-agent OSMI-1 or anti-PD-L1 antibody treatment cannot significantly increased the proportion of CD4+ T and CD8+ T cells (Fig. 7L). Production of IFN-γ and TNF-α were significantly enhanced in intratumoral CD8+ T cells in combined treatment with OSMI-1 and anti-PD-L1 antibody, but not in single-agent OSMI-1 or anti-PD-L1 treatment (Fig. 7M-N). This model is also consistent with Ogt knockout tumor-bearing mice model. In summary, OGT inhibitor OSMI-1 induces DNA damage and cytosolic DNA accumulation, activates cGAS/STING pathway. Combined OGT inhibitor with anti-PD-L1 antibody markedly suppressed tumor growth and increased CD8+ T cells and production of IFN-γ and TNF-α in tumor. These results demonstrated a pivotal role of OGT inhibition in augmenting the antitumor immune response of ICB. Given the increasing importance of immunotherapy for the management of patients with that OGT inhibitors, combined with anti-PD-L1 blockade may offer a particularly attractive strategy for the treatment of colorectal and lung cancer, which are instrumental in turning ‘cold tumors’ into ‘hot tumors’.

Discussion

As an important metabolic enzyme, OGT promotes tumorigenesis by glycosylating numerous proteins. This study found that OGT levels are elevated in both human and mouse tumors (Fig. 1). Additionally, epithelial deletion of OGT inhibits colorectal tumorigenesis in mice (Fig. 2). DNA damage and DNA repair signaling pathways play pivotal roles in maintaining genomic stability and integrity by correcting impaired DNA, which otherwise can contribute to carcinogenesis [43]. These pathways are activated in response to endogenous or exogenous DNA-damaging agents, helping cells to preserve genomic stability. HCF1 is a member of the host cell factor family, involved in regulating the cell cycle and playing regulatory roles in various transcription-related processes. This study found that OGT deficiency leads to DNA damage and cytosolic DNA accumulation (Fig. 4) by regulating HCF-1 cleavage and maturation (Fig. 5). In antitumor therapies, chemotherapy and radiotherapy induce cell death by directly or indirectly causing DNA damage, thereby increasing tumor sensitivity to cancer therapies. DNA damage-inducing therapies have proven to be immensely beneficial for cancer treatment, functioning by directly or indirectly forming DNA lesions and subsequently inhibiting cellular proliferation. Therefore, targeting DNA repair pathways may represent a promising therapeutic approach for cancer treatment. In this study, we observed that deletion of OGT leads to uncontrolled expansion of DNA damage and induces cytosolic dsDNA accumulation in tumor cells, suggesting that OGT is a promising therapeutic target for cancer treatment.

The cytosolic DNA sensing pathway has emerged as the major link between DNA damage and innate immunity, DNA damage in the nucleus results in the accumulation of cytosolic DNA, which activate the cGAS–STING pathway [44–49]. This study found that OGT deficiency activates the cGAS/STING-dependent IFN-I pathway (Fig. 3), resulting in the expression of proinflammatory cytokines (e.g., IFNB1 and ISGs) and chemokines (e.g., CXCL10) in a TBK1-IRF3-dependent manner. OGT deficiency causes DNA damage and cytosolic DNA accumulation, which triggers this response. The cGAS–STING pathway is the key cytosolic DNA sensor responsible for the type I IFN production, DC activation, and subsequent priming of CD8+ T cells against tumor-associated antigens [30, 50, 51]. Recent evidence shows that proper activation of tumor cell-intrinsic immunity or innate immune cells can enhance antitumor immunity [34, 52, 53]. In this study, cGAS-STING-IIFNB1-CXCLl10 signaling axis can provide an activated signal for epitope-specific CD8+ T cell priming by antigen-presenting cells. Ultimately, this process results in increased infiltration of tumor-infiltrating CD8+ T lymphocytes and more effective inhibition of tumor growth within the tumor microenvironment (TME) (Fig 6). However, knockout of cGAS or STING can eliminate the proliferation signal of CD8+ T cells and abolish antitumor immunity (Supplemental Fig. 11).

Immunotherapy with checkpoint-blocking antibodies targeting CTLA-4 and PD-1/PD-L1 has revolutionized cancer treatment and drastically improved the survival of individuals in the clinical treatment. Although immunotherapy has made great progress in the treatment of solid tumors, only around 20% of patients with non-small cell lung cancer (NSCLC) respond to mono-immunotherapy, and a large proportion of individuals develop resistance. Therefore, there is a need to explore novel alternative strategies and personalized immunotherapy strategies through combinations of PD-1/PD-L1 blockade with small molecular targets. This is aimed at improving sensitivity to activated antitumor immune responses in patients and addressing drug resistance [54, 55]. Here we showed that the OGT inhibitor OSMI-1 induced DNA damage and cytosolic DNA accumulation which led to activation of the cGAS-STING-TBK1-IRF3 pathway, then enhanced the innate and adaptive immune responses to tumor cells, which reversed the immunosuppressive TME by increasing CD8+ T cells infiltration. To our surprise, especially in lung cancer mice model, combination therapy with OSMI-1/PD-L1 can achieve a better antitumor effect than either monotherapy (Fig7).

In summary, our findings demonstrated deficiency in OGT mediated genomic instability and result in cytosolic dsDNA accumulation, which activating the cGAS-STING signaling pathway, increasing inflammatory cytokines, and enhancing antitumor immunity. Our study also addresses an unmet clinical need through the combination of OGT inhibition and anti-PD-L1 therapy, which may represent a promising strategy for colorectal and lung cancer therapy.